Magnet analysis

### 2024-05-14 paper/manuscript.Rmd
work on revisions round 2

Here, we propose a cognitive model of gaze following, which builds upon the notion of eye gaze as line-of-sight tracing and extends this by explicitly modeling individual differences. Our gaze model assumes participants infer the locus of someone’s attention to be where two estimated gaze vectors meet. Each of these gaze vectors results from connecting the center of the agent’s eyeball and the center of the pupil. Because the center of the eyeball is not directly observable, vector estimation happens with a degree of uncertainty. Development of gaze following corresponds to a decrease in uncertainty. Individual differences correspond to systematic differences in uncertainty.

@@

We found very clear support for our gaze model, both in children as well as adults. A strong correlation between the data mean and the gaze model estimate ($\sigma_v$) showed that the data mean is suitable to quantify individual differences: child sample *r* = `r gazemodel_corrs$r[gazemodel_corrs$corr == "gazemodel_param_kids"]`, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "gazemodel_param_kids"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "gazemodel_param_kids"]`]; adult sample *r* = `r gazemodel_corrs$r[gazemodel_corrs$corr == "gazemodel_param_adults"]`, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "gazemodel_param_adults"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "gazemodel_param_adults"]`]). The gaze model predicted a U-shaped pattern which we also observed in our data (see Figure \@ref(fig:fig3)C). The model comparison strongly favored our gaze model over the center bias model (child sample $logBF_{10}$ = `r gazemodel_comparison_kids$bf_inf_center`; adult sample $logBF_{10}$ = `r gazemodel_comparison_adults$bf_inf_center`) and the random guessing model (child sample $logBF_{10}$ = `r gazemodel_comparison_kids$bf_inf_guess`; adult sample $logBF_{10}$ = `r gazemodel_comparison_adults$bf_inf_guess`). For the child sample, we went one step further into the analysis. When correlating the observed data across all target positions with the predictions of the three models, we found a high similarity for the gaze model: *r* = `r gazemodel_corrs$r[gazemodel_corrs$corr == "gazemodel_pred_kids"]`, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "gazemodel_pred_kids"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "gazemodel_pred_kids"]`], while the correlations with the alternative models were smaller (center bias model: *r* = `r gazemodel_corrs$r[gazemodel_corrs$corr == "centermodel_pred_kids"]`, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "centermodel_pred_kids"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "centermodel_pred_kids"]`]; random guessing model: *r* = `r gazemodel_corrs$r[gazemodel_corrs$corr == "randommodel_pred_kids"]`, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "randommodel_pred_kids"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "randommodel_pred_kids"]`]). The gaze model's prior over potential target locations assumed that participants' clicks would be skewed towards the screen center. For children, we estimated the width of the prior's distribution based on the participants' age. Interestingly, we found that the differential age effects on the center bias prior were minor (intercept = `r gazemodel_prior$mode[gazemodel_prior$type == "intercept"]`, 95%HDI [`r gazemodel_prior$lci[gazemodel_prior$type == "intercept"]`; `r gazemodel_prior$uci[gazemodel_prior$type == "intercept"]`]; slope = `r gazemodel_prior$mode[gazemodel_prior$type == "slope"]`, 95%HDI [`r gazemodel_prior$lci[gazemodel_prior$type == "slope"]`; `r gazemodel_prior$uci[gazemodel_prior$type == "slope"]`]). The slope of the prior indicated that older children were only marginally less drawn towards the screen center compared to the younger children.

(ref:figlab3) **Gaze model results** (A) Developmental trajectory of the estimated model parameter. Grey dots show individual level parameter values. The black line shows the maximum a posteriori (MAP) estimate; blue lines show 1000 draws from it. The large black point with 95% HDI shows the mean of the model parameter for the adult sample. We added minimal horizontal and vertical noise to the adult individual level parameter values to avoid overplotting. (B) Correlation between estimated mode of the model parameter and data mean per individual, faceted by age group. In the child sample, color denotes age in years. Grey regression lines with 95% CI show smooth conditional means based on linear models, with *Pearson*'s correlation coefficient *r*. (C) Pattern recovery. Imprecision in target width for each target bin by age group. Model predictions in blue; data in grey. (D) Correlation between the observed data and the predictions of the three models by target position and age group (across individuals; only for the child sample).

@@

@@

Gaze following substantially correlated with the non-social vector following task, *r* = `r magnet_cormatrix$r[magnet_cormatrix$Correlation == "Tango x Magnet"]`, 95%CI [`r magnet_cormatrix$lower[magnet_cormatrix$Correlation == "Tango x Magnet"]`, `r magnet_cormatrix$upper[magnet_cormatrix$Correlation == "Tango x Magnet"]`] (see Figure \@ref(fig:fig4)B). While the tasks highly overlap in task demands, their measures shared some, but not all of their variance.

ToM abilities did not correlate with gaze following ($\rho$ = `r magnet_cormatrix$r[magnet_cormatrix$Correlation == "Tango x ToM"]`, 95%CI [`r magnet_cormatrix$lower[magnet_cormatrix$Correlation == "Tango x ToM"]`, `r magnet_cormatrix$upper[magnet_cormatrix$Correlation == "Tango x ToM"]`]) or non-social vector following ($\rho$ = `r magnet_cormatrix$r[magnet_cormatrix$Correlation == "Magnet x ToM"]`, 95%CI [`r magnet_cormatrix$lower[magnet_cormatrix$Correlation == "Magnet x ToM"]`, `r magnet_cormatrix$upper[magnet_cormatrix$Correlation == "Magnet x ToM"]`]), and the correlations did not differ from each other, Williams' test *t*(`r magnet_williamsttest_tom$parameter`) = `r round(magnet_williamsttest_tom$statistic, 2)`, *p* = `r round(magnet_williamsttest_tom$p.value, 2)`.

Interestingly, gaze following and perspective-taking correlated with each other, $\rho$ = `r magnet_cormatrix$r[magnet_cormatrix$Correlation == "Tango x Perspective-taking"]`, 95%CI [`r magnet_cormatrix$lower[magnet_cormatrix$Correlation == "Tango x Perspective-taking"]`, `r magnet_cormatrix$upper[magnet_cormatrix$Correlation == "Tango x Perspective-taking"]`]. Please note that the TANGO quantifies imprecision in gaze following. Therefore, a negative correlation suggests that imprecision in gaze following corresponds to less perspective-taking. Non-social vector following and perspective-taking did not correlate, $\rho$ = `r magnet_cormatrix$r[magnet_cormatrix$Correlation == "Magnet x Perspective-taking"]`, 95%CI [`r magnet_cormatrix$lower[magnet_cormatrix$Correlation == "Magnet x Perspective-taking"]`, `r magnet_cormatrix$upper[magnet_cormatrix$Correlation == "Magnet x Perspective-taking"]`]. However, according to the Williams' test, the two correlations did not differ significantly from each other, *t*(`r magnet_williamsttest_perspective$parameter`) = `r round(magnet_williamsttest_perspective$statistic, 2)`, *p* = `r round(magnet_williamsttest_perspective$p.value, 2)`.

@@

Gaze as displayed in the TANGO versus as in real-life social interactions differs with regard to which information is available. In the TANGO, the agent’s eyes are big and round, with uniform colors, white sclera, fully visible iris and pupils, and continuous eye movement. Eye gaze in natural social interactions often provides less information and is more ambiguous. Even though the TANGO and our gaze model are designed within a 2D world, we believe the mechanisms can be extrapolated into the 3D world. The processes of understanding gaze in daily life likely rely on the same principles as proposed in this paper. In Study 3, we presented the first evidence that this might be the case. We administered Level 2 perspective-taking tasks in which participants needed to adapt another person's frame of reference in a real-world social interaction. The correlation between this task and the TANGO speaks toward a unified mechanism behind these two visual perspective-taking tasks, regardless of the testing setup or stimulus features. Clearly, our real-life environment is visually more cluttered and diverse than the one presented in our tablet task. Here, however, additional informational sources are available to infer where others are looking; for example, body or head orientation or common ground [@bohn2018common; @moll2013primacy; @osborne-crowley2020social]. From a modeling perspective, a shared interaction history or diverse desires might be represented as non-uniform prior distributions over locations in the visual scenery. This way, our gaze model could be expanded to include more complex processes like mental state reasoning.

>>>>

Here, we propose a cognitive model of gaze following, which builds upon the notion of eye gaze as line-of-sight tracing and extends this by explicitly modeling individual differences. Our gaze model assumes participants infer the locus of someone’s attention to be where two estimated gaze vectors meet. Each of these gaze vectors results from connecting the center of the agent’s eyeball and the center of the pupil. Because the center of the eyeball is not directly observable at that moment, vector estimation happens with a degree of uncertainty. Development of gaze following corresponds to a decrease in uncertainty. Individual differences correspond to systematic differences in uncertainty.

@@

We found very clear support for our gaze model, both in children as well as adults. A strong correlation between the data mean and the gaze model estimate ($\sigma_v$) showed that the data mean is suitable to quantify individual differences: child sample *r*(`r gazemodel_corrs$n[gazemodel_corrs$corr == "gazemodel_param_kids"]`) = `r gazemodel_corrs$r[gazemodel_corrs$corr == "gazemodel_param_kids"]`,  *p* < .001, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "gazemodel_param_kids"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "gazemodel_param_kids"]`]; adult sample *r*(`r gazemodel_corrs$n[gazemodel_corrs$corr == "gazemodel_param_adults"]`) = `r gazemodel_corrs$r[gazemodel_corrs$corr == "gazemodel_param_adults"]`,  *p* < .001, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "gazemodel_param_adults"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "gazemodel_param_adults"]`]. The gaze model predicted a U-shaped pattern which we also observed in our data (see Figure \@ref(fig:fig3)C). The model comparison strongly favored our gaze model over the center bias model (child sample $logBF_{10}$ = `r gazemodel_comparison_kids$bf_inf_center`; adult sample $logBF_{10}$ = `r gazemodel_comparison_adults$bf_inf_center`) and the random guessing model (child sample $logBF_{10}$ = `r gazemodel_comparison_kids$bf_inf_guess`; adult sample $logBF_{10}$ = `r gazemodel_comparison_adults$bf_inf_guess`). For the child sample, we went one step further into the analysis. When correlating the observed data across all target positions with the predictions of the three models, we found a high similarity for the gaze model: *r*(`r gazemodel_corrs$n[gazemodel_corrs$corr == "gazemodel_pred_kids"]`) = `r gazemodel_corrs$r[gazemodel_corrs$corr == "gazemodel_pred_kids"]`,  *p* < .001, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "gazemodel_pred_kids"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "gazemodel_pred_kids"]`], while the correlations with the alternative models were smaller (center bias model: *r*(`r gazemodel_corrs$n[gazemodel_corrs$corr == "centermodel_pred_kids"]`) = `r gazemodel_corrs$r[gazemodel_corrs$corr == "centermodel_pred_kids"]`,  *p* < .001, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "centermodel_pred_kids"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "centermodel_pred_kids"]`]; random guessing model: *r*(`r gazemodel_corrs$n[gazemodel_corrs$corr == "randommodel_pred_kids"]`) = `r gazemodel_corrs$r[gazemodel_corrs$corr == "randommodel_pred_kids"]`,  *p* < .001, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "randommodel_pred_kids"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "randommodel_pred_kids"]`]). The gaze model's prior over potential target locations assumed that participants' clicks would be skewed towards the screen center. For children, we estimated the width of the prior's distribution based on the participants' age. Interestingly, we found that the differential age effects on the center bias prior were minor (intercept = `r gazemodel_prior$mode[gazemodel_prior$type == "intercept"]`, 95% high-density interval (HDI) [`r gazemodel_prior$lci[gazemodel_prior$type == "intercept"]`; `r gazemodel_prior$uci[gazemodel_prior$type == "intercept"]`]; slope = `r gazemodel_prior$mode[gazemodel_prior$type == "slope"]`, 95%HDI [`r gazemodel_prior$lci[gazemodel_prior$type == "slope"]`; `r gazemodel_prior$uci[gazemodel_prior$type == "slope"]`]). The slope of the prior indicated that older children were only marginally less drawn towards the screen center compared to the younger children.

(ref:figlab3) **Gaze model results** (A) Developmental trajectory of th±e estimated model parameter. Grey dots show individual level parameter values. The black line shows the maximum a posteriori (MAP) estimate; blue lines show 1000 draws from it. The large black point with 95% HDI shows the mean of the model parameter for the adult sample. We added minimal horizontal and vertical noise to the adult individual level parameter values to avoid overplotting. (B) Correlation between estimated mode of the model parameter and data mean per individual, faceted by age group. In the child sample, color denotes age in years. Grey regression lines with 95% CI show smooth conditional means based on linear models, with *Pearson*'s correlation coefficient *r*. (C) Pattern recovery. Imprecision in target width for each target bin by age group. Model predictions in blue; data in grey. (D) Correlation between the observed data and the predictions of the three models by target position and age group (across individuals; only for the child sample).

@@

Previous models in the social- and non-social cognitive domain exist that rely on vector estimations (see @madl2015computational for a review of computational models for spatial memory). For example, @brom2012computational defined egocentric self-to-object vectors and allocentric object-to-object vectors where corresponding uncertainties due to memory and motor errors were modelled by vector weights and Gaussian noise. The counterfactual simulation model [@gerstenberg2021counterfactuala, @gerstenberg2021counterfactual, @gerstenberg2012noisy, @gerstenberg2014counterfactual] used vector calculations to model causal reasoning, for example, in estimating billiard ball trajectories and their collisions. Gaussian noise was added to the ball’s movement vector to represent people’s prior expectations, uncertainties, and expertise (the stronger the expectations, the less noise is added to the vector).

In the social domain, there are additional examples of following vectors in space. Recent work on pointing gestures introduced the idea of a touch-line vector [@omadagain2019origin], see @li2023understanding for computational modeling. According to this hypothesis, humans directed their fingers not to construct an "arrow" pointing at the target, but rather as if they aimed to touch it directly. Consequently, the touch line is a vector that connects the eye and the pointed fingertip, and marks the object that the person's fingertip seems closest to touching from her perspective [@omadagain2019origin]. This supports the idea that vector estimation is relevant for social-cognitive abilities.

@@

Gaze following substantially correlated with the non-social vector following task, *r*(`r magnet_cormatrix$n[magnet_cormatrix$Correlation == "Tango x Magnet"]`) = `r magnet_cormatrix$r[magnet_cormatrix$Correlation == "Tango x Magnet"]`, *p* < .001, 95%CI [`r magnet_cormatrix$lower[magnet_cormatrix$Correlation == "Tango x Magnet"]`, `r magnet_cormatrix$upper[magnet_cormatrix$Correlation == "Tango x Magnet"]`] (see Figure \@ref(fig:fig4)B). While the tasks highly overlap in task demands, their measures shared some, but not all of their variance.

ToM abilities did not correlate with gaze following ($\rho$(`r magnet_cormatrix$n[magnet_cormatrix$Correlation == "Tango x ToM"]`) = `r magnet_cormatrix$r[magnet_cormatrix$Correlation == "Tango x ToM"]`, *p* = `r magnet_cormatrix$p[magnet_cormatrix$Correlation == "Tango x ToM"]`, 95%CI [`r magnet_cormatrix$lower[magnet_cormatrix$Correlation == "Tango x ToM"]`, `r magnet_cormatrix$upper[magnet_cormatrix$Correlation == "Tango x ToM"]`]) or non-social vector following ($\rho$(`r magnet_cormatrix$n[magnet_cormatrix$Correlation == "Magnet x ToM"]`) = `r magnet_cormatrix$r[magnet_cormatrix$Correlation == "Magnet x ToM"]`,  *p* = `r magnet_cormatrix$p[magnet_cormatrix$Correlation == "Magnet x ToM"]`, 95%CI [`r magnet_cormatrix$lower[magnet_cormatrix$Correlation == "Magnet x ToM"]`, `r magnet_cormatrix$upper[magnet_cormatrix$Correlation == "Magnet x ToM"]`]), and the correlations did not differ from each other, Williams' test *t*(`r magnet_williamsttest_tom$parameter`) = `r round(magnet_williamsttest_tom$statistic, 2)`, *p* = `r round(magnet_williamsttest_tom$p.value, 3)`.

Interestingly, gaze following and perspective-taking correlated with each other, $\rho$(`r magnet_cormatrix$n[magnet_cormatrix$Correlation == "Tango x Perspective-taking"]`) = `r magnet_cormatrix$r[magnet_cormatrix$Correlation == "Tango x Perspective-taking"]`, *p* = `r magnet_cormatrix$p[magnet_cormatrix$Correlation == "Tango x Perspective-taking"]`, 95%CI [`r magnet_cormatrix$lower[magnet_cormatrix$Correlation == "Tango x Perspective-taking"]`, `r magnet_cormatrix$upper[magnet_cormatrix$Correlation == "Tango x Perspective-taking"]`]. Please note that the TANGO quantifies imprecision in gaze following. Therefore, a negative correlation suggests that imprecision in gaze following corresponds to less perspective-taking. Non-social vector following and perspective-taking did not correlate, $\rho$(`r magnet_cormatrix$n[magnet_cormatrix$Correlation == "Magnet x Perspective-taking"]`) = `r magnet_cormatrix$r[magnet_cormatrix$Correlation == "Magnet x Perspective-taking"]`, *p* = `r magnet_cormatrix$p[magnet_cormatrix$Correlation == "Magnet x Perspective-taking"]`, 95%CI [`r magnet_cormatrix$lower[magnet_cormatrix$Correlation == "Magnet x Perspective-taking"]`, `r magnet_cormatrix$upper[magnet_cormatrix$Correlation == "Magnet x Perspective-taking"]`]. However, according to the Williams' test, the two correlations did not differ significantly from each other, *t*(`r magnet_williamsttest_perspective$parameter`) = `r round(magnet_williamsttest_perspective$statistic, 2)`, *p* = `r round(magnet_williamsttest_perspective$p.value, 3)`.

@@

Gaze as displayed in the TANGO versus as in real-life social interactions differs with regard to which information is available. In the TANGO, the agent’s eyes are big and round, with uniform colors, white sclera, fully visible iris and pupils, and continuous eye movement. The start position of the pupil is the center of the agent's eyeball -- which, later on, needs to be estimated. Eye gaze in natural social interactions often provides less information and is more ambiguous. Even though the TANGO and our gaze model are designed within a 2D world, we believe the mechanisms can be extrapolated into the 3D world. The processes of understanding gaze in daily life likely rely on the same principles as proposed in this paper. In Study 3, we presented the first evidence that this might be the case. We administered Level 2 perspective-taking tasks in which participants needed to adapt another person's frame of reference in a real-world social interaction. The correlation between this task and the TANGO speaks toward a unified mechanism behind these two visual perspective-taking tasks, regardless of the testing setup or stimulus features. Clearly, our real-life environment is visually more cluttered and diverse than the one presented in our tablet task. Here, however, additional informational sources are available to infer where others are looking; for example, body or head orientation or common ground [@bohn2018common; @moll2013primacy; @osborne-crowley2020social]. From a modeling perspective, a shared interaction history or diverse desires might be represented as non-uniform prior distributions over locations in the visual scenery. This way, our gaze model could be expanded to include more complex processes like mental state reasoning.

### 2024-05-14 stats/gazemodel.Rmd
work on revisions round 2

corrs <- rbind(corr_gazemodel_param_children$ci, corr_gazemodel_param_adults$ci, corr_gazemodel_pred$ci, corr_centermodel_pred$ci, corr_randommodel_pred$ci) %>% 
  rownames_to_column(var = "corr") %>% 
  mutate(
    corr = recode(
            corr,
            `mdl_m-dt_mn` = "gazemodel_param_kids", 
            `mdl_m-dt_mn1` = "gazemodel_param_adults", 
            `mdl_m-dt_mn2` = "gazemodel_pred_kids", 
            `mdl_m-dt_mn3` = "centermodel_pred_kids", 
            `mdl_m-dt_mn4` = "randommodel_pred_kids", 
           )
  ) %>% 
  mutate(across(where(is.numeric), round, 2))

>>>>

corrs <- rbind(
  cbind(corr_gazemodel_param_children$ci, n = corr_gazemodel_param_children$n), 
  cbind(corr_gazemodel_param_adults$ci, n = corr_gazemodel_param_adults$n), 
  cbind(corr_gazemodel_pred$ci, n = corr_gazemodel_pred$n), 
  cbind(corr_centermodel_pred$ci, n = corr_centermodel_pred$n), 
  cbind(corr_randommodel_pred$ci, n = corr_randommodel_pred$n)
  ) %>% 
  rownames_to_column(var = "corr") %>% 
  mutate(
    corr = recode(
            corr,
            `mdl_m-dt_mn` = "gazemodel_param_kids", 
            `mdl_m-dt_mn1` = "gazemodel_param_adults", 
            `mdl_m-dt_mn2` = "gazemodel_pred_kids", 
            `mdl_m-dt_mn3` = "centermodel_pred_kids", 
            `mdl_m-dt_mn4` = "randommodel_pred_kids", 
           ), 
  ) %>% 
  mutate(across(2:4, ~round(.x, 2)))

### 2024-05-14 stats/magnet.Rmd
work on revisions round 2

cor_matrix <- rbind(tango_magnet$ci, tango_tom$ci, magnet_tom$ci) %>% 
  mutate(Correlation = rownames(.)) %>%
  as.tibble()

cor_matrix <- cor_matrix %>%
  # remove tom - perspective correlations
  filter(Correlation != "tm_gg-prsp_1") %>%
  mutate(
    Correlation = recode(Correlation,
                  `tng_m-mgnt_` = "Tango x Magnet", 
                  `tng_m-tm_gg` = "Tango x ToM",
                  `tng_m-prsp_` = "Tango x Perspective-taking",
                  `mgnt_-tm_gg` = "Magnet x ToM",
                  `mgnt_-prsp_` = "Magnet x Perspective-taking", 
                  `tm_gg-prsp_` = "ToM x Perspective-taking")
  ) %>% 
  mutate(across(where(is.numeric), round, 2)) %>% 
  select(Correlation, r, lower, upper, p)

>>>>

cor_matrix <- rbind(
  cbind(tango_magnet$ci, n = tango_magnet$n),
  cbind(tango_tom$ci, n = tango_tom$n), 
  cbind(magnet_tom$ci, n = magnet_tom$n)
  ) %>% 
  mutate(Correlation = rownames(.)) %>%
  as_tibble()

cor_matrix <- cor_matrix %>%
  # remove tom - perspective correlations
  filter(Correlation != "tm_gg-prsp_1") %>%
  mutate(
    Correlation = recode(Correlation,
                  `tng_m-mgnt_` = "Tango x Magnet", 
                  `tng_m-tm_gg` = "Tango x ToM",
                  `tng_m-prsp_` = "Tango x Perspective-taking",
                  `mgnt_-tm_gg` = "Magnet x ToM",
                  `mgnt_-prsp_` = "Magnet x Perspective-taking", 
                  `tm_gg-prsp_` = "ToM x Perspective-taking"), 
    p = round(p, 3)
  ) %>% 
  mutate(across(1:3, ~round(.x, 2))) %>%
  select(Correlation, r, lower, upper, p, n)
